---
title: "Regression Models - Course Project"
author: "Philip Mateescu"
date: "December 26, 2015"
output: html_document
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(GGally)
library(ggfortify)
library(datasets)
library(dplyr)
library(knitr)
library(car)
arsq <- function(model) summary(model)$adj.r.squared
    
```

# Impact of Transmission Type of Automobile Mileage

## Synopsis/Executive Summary

Fuel efficiency, expressed in miles-per-gallon, or *mpg* in short, is a common selection criteris when comparing cars.

Using a dataset extracted from the 1974 *Motor Trend US* magazine, 
and comprising of fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973–74 models),
we will explore the relationship between automatic or manual transmission types and the fuel consumption to answer 
the following questions:

- “Is an automatic or manual transmission better for MPG”
- "Quantify the MPG difference between automatic and manual transmissions"


Using simple linear regression analyses, we have determined that there is a significant difference between 
automatic and manual transmissions in cars. The cars equiped with manual transmission obtain an average of 
7.97 more miles per gallon when compared to cars equiped with automatic transmission, when accounting 
for all other factors impacting fuel consumption.


## Exploratory Data Analysis

In order to make better use of the `mtcars` dataset I have changed the type of some of the columns
to better reflect their underlying types, for example changing the `cyl` column, 
the number of cylinders, from a `numeric` to a `factor`, to indicate that this is *not*
a continuous measurement and changing the factors of the `am` column from `0, 1` to `A, M`.  
See **Appendix - Table 1** for all changes made and **Appendix - Table 2** for a sample dataset.


```{r echo=FALSE}
oldtypes <- sapply(mtcars, class)
mtcars$cyl <- as.factor(mtcars$cyl)
mtcars$am <- as.factor(mtcars$am)
levels(mtcars$am) <- c('A', 'M')
mtcars$gear <- as.factor(mtcars$gear)
mtcars$carb <- as.factor(mtcars$carb)
newtypes <- sapply(mtcars, class)
old.to.new.columns <- data.frame(Old.Types=oldtypes, New.Types=newtypes)
```
```{r echo=FALSE}
a.avg <-  mean(mtcars$mpg[which(mtcars$am == 'A')])
a.sd <- sd(mtcars$mpg[which(mtcars$am == 'A')])
m.avg <- mean(mtcars$mpg[which(mtcars$am == 'M')])
m.sd <- sd(mtcars$mpg[which(mtcars$am == 'M')])
```

**Appendix - Table 3**  shows there is a difference between the average mpg of automatic cars --
``r a.avg``mpg, with a standard deviation of 
``r a.sd`` -- and the **higher** average mpg 
of cars equipped with manual transmission, at ``r m.avg``mpg, 
with a standard deviation of ``r m.sd``.  
**Appendix - Figure 1** shows a box-plot of the transmission/mpg relationship.

Fitting a linear model with `am` as the regressor and `mpg` as the outcome,
shows there is a significant relationship between the two (*p-value* < 0.001).  
See **Appendix - Table 4 - Summary of base linear model**.

```{r}
model.base <- lm(mpg ~ am, mtcars)
```

However, an *Adjusted R-squared* of 0.34 indicates only about a third 
of the variation in mileage can be explained by the choice of transmission.


## Statistical Inference

Under the assumption that the `mpg` has a normal distribution, w
we  the null hypothesis, *H0*, as: manual and automatic transmission 
come from the same population. 

```{r}
h0 <- t.test(mpg ~ am, data = mtcars)
```

Using an unpaired, two-sided T-test at 95% confidence level, yields a
p-value of ``r h0$p.value``. Since the *p-value* is less than 0.05,
we reject our null hypothesis, thus
the automatic and manual transmissions come from different populations.

## Model Selection

We will use the [Akaike information criterion](https://en.wikipedia.org/wiki/Akaike_information_criterion)
and the step function to choose the model best fitted to explain the variation in mpg. 

```{r}
model.best <- step(lm(mpg ~ ., mtcars), k=log(nrow(mtcars)), trace = 0)
```

According to **Appendix - Table 5 - Summary of best linear model**, 
the model using `wt` - weight, `qsec` - quarter-mile time, and `am` - transmission type,
explains 83% percent of the variance of the `mpg` variable., and all variables are 
significant at the 0.05 level or smaller.

However, in consulting with our imaginary engine experts, we have learned that 
the quarter-mile time is likely inversely proportional to the horsepower of the car (when weight is constant).

Horsepower itself is a complex relationship of engine configuration (size, number of cylinders, engine geometry, number of carburetors)
and gearing - `gear`. All there variables 
are likely confounding for the quarter-mile time. Let's examine a model that regresses on horsepower 
and compare it to a model that uses the individual variables.

```{r}
model.amwt <- lm(mpg ~ am + wt, mtcars)
model.hp <- update(model.amwt, mpg ~ am + wt + hp)
model.engine <- update(model.amwt, mpg ~ am + wt + disp + gear + cyl + carb + vs)
df.comp1 <- data.frame(
    arsq(model.amwt), 
    arsq(model.hp), 
    arsq(model.engine))
kable(df.comp1, col.names = c('am+wt', 'am+wt+hp', 'am+wt+disp+gear+cyl+carb+vs'), caption = 'Adjusted R-squared values for 3 models')
```

We notice that while adding the engine configuration explains the variation better than weight + transmission type alone, 
adding the horsepower to the `am + wt` model explains 82% of the variance, close to the best model chosen by 
the step function, but considerably easier to explain even to novice readers.

Going back to our cars experts with our findings, we found them keen to point out that there is an interaction
between the transmission type and the weight of the car, given that automatic transmissions tend to simply weigh more.  
Armed with this new knowledge, let's consider a model where the `wt` and `am` interact.

```{r}
model.final <- lm(mpg ~ am + wt + am:wt + hp, mtcars)
```

Surprisingly (or perhaps not), a collaboration between the statistics student and the car experts
produce a model that explains the variance even better than the model selected by the step function:
a whooping **85%** and all variables statistically significant at level 0.05 or lower! 
See **Appendix - Table 6 - Summary of final linear model**.

(Note: we'd be disingenuous if we didn't point out that if we take `step`'s best model and include the `am:wt` interaction we
get an even better ``r arsq(lm(mpg ~ am + wt + qsec + am:wt, mtcars))`` adjusted R-squared).

This model shows that when weight and horsepower remain constant, cars with manual transmission 
get `11.55 + (-3.58) =` ``r 11.55-3.58`` more miles per gallong than cars equiped with automatic transmission.

### Residuals and Diagnostics

In this section, we'll perform a few diagnostics of our model, examine the residuals and leverage variables
in order to detect any potential problems with our model.

We start with a diagnostic plot of our model, shows in the **Appendix - Figure 2**. From these four chart 
can observe that:

- There is no consistent pattern when examining *Fitted Values vs Residuals*, thus supporting the independence of our chosed predictors;
- The *Normal Q-Q* plot indicates the standardized residuals are normally distributed and close to out fitted line;
- The *Scale-Location* chart shows scatter, thus confirming constant variance,
- Finally, the *Residuals vs Leverage* chart, shows that while we have outliest, none have considerable leverage (all within the 0-0.5 band).

The `hatvalues` function gives us the leverage points in the model. The top 3 points are:
``` {r}
leverage <- hatvalues(model.final)
tail(sort(leverage), n = 3)
```

The top 3 most influential cars can be found using the `dfbetas` function:
```{r}
influence <- dfbetas(model.final)
tail(sort(influence[,2]), n = 3)
```

\pagebreak

# Appendix

## Tables

#### Table 1 - Changes in variable types
```{r}
kable(old.to.new.columns, caption = '')
```

#### Table 2 - Sample data from *mtcars*
```{r}
print(mtcars[2:5,])
```

#### Table 3 - Averages and standard deviations of fuel consumptions for automatic and manual cars

```{r}
avgsd <- data.frame(Automatic = c(a.avg, a.sd), Manual = c(m.avg, m.sd))
row.names(avgsd) <- c('Average mpg', 'Std dev')
kable(avgsd, row.names = TRUE)
```

#### Table 4 - Summary of base linear model
```{r}
summary(model.base)
```
#### Table 5 - Summary of best linear model
```{r}
summary(model.best)
```

#### Table 6 - Summary of final linear model
```{r}
summary(model.final)
```


## Charts

#### Figure 1 - Transmission Type vs MPG

.

```{r}
ggplot(mtcars, aes(am, mpg, col = am)) + geom_boxplot() + geom_jitter(aes(col=am)) +
    labs(x='Transmission Type', y = 'mpg') + 
    theme(legend.position='none')
 
```

#### Figure 2 - Diagnostic Plots for model *mpg ~ am + wt + am:wt + hp*

.

```{r}
# using ggplot and ggfortiy to create a diagnostic plot similar to the one plot.lm produces
# but prettier
# source: http://rpubs.com/sinhrks/plot_lm
autoplot(model.final, data = mtcars, colour = 'am', label.size=3, main='Model mpg ~ am + wt + am:wt + hp') + 
    theme(legend.position='none')
```
**Legend**: red dots - Automatic transmission, teal dots - Manual transmission

